---
title: "final_ENVS193DS"
author: "Ian Del Tredici"
date: "03-12-2026"
format:
  html:
    toc: TRUE
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

[Github Repo](https://github.com/iandeltredici/ENVS-193DS_winter-2026_final)

```{r}
#| label: load in packages and data
library(tidyverse) #load packages
library(janitor)
library(here)
library(readxl)
library(ggeffects)
library(showtext)
font_add_google("Noto Serif Georgian")
showtext_auto()
reading_data <- read.csv(here("data", "reading_data_final.csv")) #load reading data
kelp <- read.csv(here("data", "kelp_data_LTER.csv"))
mopl <- read.csv(here("data", "MOPL_nest-site_selection_Duchardt_et_al._2019.csv"))
```

# Problem 1. Research Writing
## a. Transparent statistical methods
In part 1 they used a simple linear regression model to determine if precipitation predicts flooded wetland area. In part 2 they used Kruskal-Wallis test to compare medians between two variables with one variable (water year classification) having multiple levels.

## b. Figure needed
Five boxplots with water year classification on the x-axis and flooded wetland are would be the best method for displaying their test. Boxplots are the perfect visualization because they display the median and allow comparison between the five groups.

## c. Suggestions for rewriting
Part 1: Precipitation did not significantly predict flooded wetland area (simple linear regression, F(df model, df error) = test statistic, R² = effect size, p = 0.11, α = significance level). Therefore we fail to reject the null hypothesis that precipitation does not predict flood wetland area.

Part 2: There was no significant difference in median flooded wetland areas between water year classifications (wet, above normal, below normal, dry, critical drought) (Kruskal-Wallis test, χ²(4) = test statistic, p = 0.12, α = significance level).

## d. Interpretation
The number of water control structures in the area, a discrete variable, could influence the flooded area of a wetland because it directly relates to how much influence wetland managers have when controlling flooding. A higher number of water control structures would give more precision, allowing wetland managers to dial in the amount of they allow to flow through, thereby dramatically affecting how much wetland is flooded.

# Problem 2. Data visualization
## a. Cleaning and summarizing
```{r}
#| label: cleaning and summarizing
#| message: false
kelp_clean <- kelp |> 
  clean_names() |>  # function 3
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> #function 6
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |>  #function 1 
  mutate(site_name = case_when( #function 8
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria"
  )) |>
  mutate(site_name = as_factor(site_name), #function 2
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria"))|> 
  group_by(site_name, year) |>  #function 5
  summarize(mean_fronds = mean(fronds)) |> #function 7
  ungroup() #function 4


```

## b. Understanding the data
Data collected on 2000-09-22 in transect 6 and quad 40 was removed because the data sheet went missing but the value of the fronds in that quad in that transect is -99999. This information was found on row 73 of the "kelp" data frame under the notes column that reported "missing datasheet".

## c. Visualize data
```{r}
#| label: line plot
#| message: false
ggplot(data = kelp_clean, 
       aes( x = year,
            y = mean_fronds,
            color = site_name)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = c(
    "Naples" = "forestgreen",
    "Isla Vista" = "darkseagreen2",
    "Carpinteria" = "aquamarine3")) +
  labs(title = "Sites vary in mean kelp fronds per year",
       x = "Year",
       y = "Mean kelp fronds per year",
       color = "Site") +
  theme_minimal(base_family = "Noto Serif Georgian") +
  theme(plot.title = element_text(hjust = 0.5),
      panel.border = element_blank())
slice_sample(kelp_clean, n=5)
str(kelp_clean)
```
## d. Interpretation
Isla Vista is the site with the most variable mean kelp fronds per year because you can see the line connecting the data points stretching all the way up to nearly 40 kelp fronds and then back down to below 10. Visually, you can see the data line for Isla Vista dance all over the graph, the data points are farther apart and less consistent than the other two sites.
Carpinteria is the site with the least variable mean kelp fronds per year, visually identifiable because the data points are more consistent and the overall trend of the line does not exhibit massive swings in kelp fronds. The line remains just below or around 10 kelp fronds per year for nearly 25 years.

# Problem 3. Data analysis
```{r}
#| label: cleaning and wrangling
#| include: false
mopl_v2 <- mopl |> 
  clean_names() |> 
  select(edgedist, id, vo_nest)
```
